Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-03-04 paper/main.tex
section 6 polished

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA can
be obtained for example by translating LTL formulas by
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf}),
\texttt{ltl3tela}~\cite{major.19.atva},
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav},
or \texttt{ltl2tgba} which is a part of the Spot library~\cite{duret.16.atva2}.
There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

We have implemented our reduction technique in a tool called
\telatko. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
ltl3tela~\cite{major.19.atva}, Rabinizer~4~\cite{kretinsky.18.cav},
and Spot~\cite{duret.16.atva2}.  While the reduction is relatively
modest for TELA produced by \texttt{ltl3tela} and Spot,
it is substantial for automata produced by tools of the Owl library.

\paragraph*{Related results}
There is a simple technique~\cite{babiak.13.spin} reducing the number
of acceptance marks in \emph{transition-based genealised Büchi
  automata (TGBA)} without changing its structure. We are not aware
about any existing research aimed at simplification of acceptance
formulas of TELA or reduction of the number of its acceptance marks
without increasing the number of states. There exists only a SAT-based
approach that transforms a deterministic TELA automaton to an
equivalent automaton with a given acceptance condition and a given
number of states~\cite{baarir.15.lpar} (if such an automaton
exists). Further, there are some SAT-based approaches aimed to reduce
the number of states of automata over infinite words. More preceisely,
there are reductions designed for nondeterministic Büchi
automata~\cite{ehlers.10.spin}, deterministic Büchi
automata~\cite{ehlers.10.sat}, and deterministic generalized Büchi
automata~\cite{baarir.14.forte}. Note that all these techniques are
usually very slow and their authors typically suggest to use them only
for specific purposes like looking for cases where some automata
construction can be improved.

@@

We have used the tool \texttt{genltl} from Spot to generate LTL formulas
from literature and from scalable patterns. Using the values from 1 to 5 for
the parametrized formulas, we have generated 407 unique LTL formulas. These
have been translated to deterministic or nondeterministic TELA using four
state-of-the-art translators \texttt{ltl2tgba} (from the Spot library),
\texttt{ltl3tela}, \texttt{ltl2dela} (known as Delag), and \texttt{ltl2dgra}
(known as Rabinizer 4). The generated automata sets contain less than 407
automata as not all of the formulas are successfully translated due to timeouts.
We have also removed automata with 0 or 1 acceptance marks as there is little
point reducing these.
Table \ref{tab:ltltranslators} shows the exact versions, the homepage of the
used tools and the number of automata produced by the given tool from
the LTL formulas after filtering.

@@

For each automata set (that is, the output of individual LTL translator), we
have run \telatko with various reduction levels, always with the incremental
solving turned on. We have again filtered out the automata for which any run
of \telatko timed out after 120 seconds. We do this to make sure the numbers
of acceptance marks can be meaningfully compared between different reduction
levels of the same automata set.

Table~\ref{tab:cumm-data} shows the number of marks in the input sets and after
each reduction level, together with the reduction ratio and time spent by the
individual reduction level. The columns \textit{solver response} and
\textit{solver timeout} show the number of automata for which the last query to
QBF solver has been successfully answered, or the solver timed out within a 30
seconds limit to a single query, respectively. An unanswered QBF formula (i.e.
the solver timed out) is considered unsatisfiable.

\begin{table}[h]
\caption{LTL to automata translators used for the experimental evaluation
and the number of successfully translated formulas.}
\label{tab:ltltranslators}
\centering
\begin{tabular}{lr@{~~~}lr}
\toprule
tool        & version&homepage & automata \\
\midrule
\texttt{ltl2dela} & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 162\\
\texttt{ltl2dgra} & 21.0 & & 274\\
\texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 89\\
\texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 108\\
\bottomrule
\end{tabular}
\end{table}

@@

Table~\ref{tab:heatmap} shows the number of automata coming from all four sets
for every input and output number of acceptance marks. The results of the
multi-level reduction technique are presented. The table indicates that in most
cases only 1 or 2 marks can be saved. However, some of the automata with
more than 7 acceptance marks could be, in theory, probably reduced more as
the query to the QBF solver mostly remains unanswered in the time limit.
Moreover, some of the results are promising, e.g. in 17 cases, we have
successfully reduced 8 or more acceptance marks to only 4.

@@

Figure~\ref{fig:quantile} presents the time spent by reducing the individual
automata sets. The chart shows a pleasing finding that for every set, most
automata are reduced in under a second and the high cummulative running times
are caused by few complicated automata.

@@

Pridat do experimentu nejaky graf, ktery ukaze, jak rychle se nektere
automaty redukovaly na minimum? Nebo spis ukazat vliv timeoutu na
redukce?  (zvysovat timeout od 1s do 50s a vzdy ukazat miru
redukce). Ty redukce ukazovat nejak jinak nez jen celkovym cislem
(boxplot? Scatterplot?).

Zkusit upravit redukci tak, aby vysledkem bylo TGBA a zkusit dale
redukovat akceptacni znacky TGBA automatu redukovanych Spotem?

Chceme ukazovat, ze redukce znacek muze nekdy umoznit dalsi redukci stavu?

>>>>

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
Spot~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

We have implemented our reduction technique in a tool called
\telatko. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
ltl3tela~\cite{major.19.atva}, Rabinizer~4~\cite{kretinsky.18.cav},
and Spot~\cite{duret.16.atva2}.  While the reduction is relatively
modest for TELA produced by ltl3tela and Spot, it is substantial for
automata produced by Rabinizer~4. \todo{A jak vyjde Delag? Fixnout i
  Abstract.}

\paragraph*{Related results}
There is a simple technique~\cite{babiak.13.spin} reducing the number
of acceptance marks in \emph{transition-based genealised Büchi
  automata (TGBA)} without changing its structure. We are not aware
about any existing research aimed at simplification of acceptance
formulas of TELA or reduction of the number of its acceptance marks
without increasing the number of states. There exists only a SAT-based
approach that transforms a deterministic TELA automaton to an
equivalent automaton with a given acceptance condition and a given
number of states~\cite{baarir.15.lpar} (if such an automaton
exists). Further, there are some SAT-based approaches aimed to reduce
the number of states of automata over infinite words. More preceisely,
there are reductions designed for nondeterministic Büchi
automata~\cite{ehlers.10.spin}, deterministic Büchi
automata~\cite{ehlers.10.sat}, and deterministic generalized Büchi
automata~\cite{baarir.14.forte}. Note that these techniques are
usually very slow and their authors typically suggest to use them only
for specific purposes like looking for cases where some automata
construction can be improved.

@@

We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 5. Altogether, we have
generated 407 unique LTL formulas. These have been translated to
deterministic TELA by two state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva} and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2}) and
\texttt{ltl3tela}~\cite{major.19.atva}. Each of the four automata sets
produced by individual tools contain less than 407 automata as not all
of the formulas are successfully translated due to timeout of 60
seconds. Further, we have removed automata with 0 or 1 acceptance
marks as there is little point reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of used translators,
the homepage of the used tools and the number of automata produced by
the given tool from the LTL formulas after filtering.

@@

For each automata set (that is, the output of an individual LTL
translator), we have applied all single-level reductions and the
multi-level reduction, always with incremental approach and without
reduction of the number of clauses. The timeout for each QBF query was
set to 30 seconds. We have again filtered out the automata for which
any of the four \telatko runs timed out after 120 seconds. We do this
to make sure the numbers of acceptance marks can be meaningfully
compared between different reduction levels of the same automata set.

Table~\ref{tab:cumm-data} shows the cummulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
columns \textit{solver response} and \textit{solver timeout} show the
number of automata for which the last query to QBF solver has been
successfully answered, or the solver timed out within a 30 seconds
limit to a single query, respectively. The timeout of the last QBF
query implies that the automaton could be potentially further reduced
by the reduction with a longer time limit.

\begin{table}[b]
\caption{LTL to automata translators used for the experimental evaluation
and the number of successfully translated formulas.}
\label{tab:ltltranslators}
\centering
\begin{tabular}{lr@{~~~}lr}
\toprule
tool        & version&homepage & automata \\
\midrule
\texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 89\\
\texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 108\\
\texttt{ltl2dela} & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 162\\
\texttt{ltl2dgra} & 21.0 & & 274\\
\bottomrule
\end{tabular}
\end{table}

@@

Table~\ref{tab:heatmap} shows the number of automata coming from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2 marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 17 cases, we have successfully reduced 8 or more
acceptance marks to only 4. 
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

@@

Figure~\ref{fig:quantile} presents the time spent by multi-level
reduction of automata in individual automata sets. The chart shows a
pleasing finding that for every set, most automata are reduced in
under a second and the high cummulative running times are caused by
few complicated automata.

@@

% Pridat do experimentu nejaky graf, ktery ukaze, jak rychle se nektere
% automaty redukovaly na minimum? Nebo spis ukazat vliv timeoutu na
% redukce?  (zvysovat timeout od 1s do 50s a vzdy ukazat miru
% redukce). Ty redukce ukazovat nejak jinak nez jen celkovym cislem
% (boxplot? Scatterplot?).

% Zkusit upravit redukci tak, aby vysledkem bylo TGBA a zkusit dale
% redukovat akceptacni znacky TGBA automatu redukovanych Spotem?

% Chceme ukazovat, ze redukce znacek muze nekdy umoznit dalsi redukci stavu?
